### 2021-07-05 rad-icfp-2021/generalisation.tex
writing generalisation section

\subsection{Lifting the restriction on map and fold}
\label{sub:Lifting the restriction on map and fold}

TODO: not sure about this, and my derivative for reduce/foldl is currently a bit wrong

Note that we recover !zip A B! as !map2 (fun x,y -> (x,y)) A B)!. 
We can lift the restriction on the function of !foldl! by changing a part in $\invUNFSymbol$ of !foldl!.
\begin{center}
\begin{tabular}{r c l}
    !Y((x1,...,xm),...)! & -> &  !Y((x1,...,xm) + map2! \\
        && !(fun z1 z2 -> let E = map2 (fun x,y -> z2(x,y)) A e3 in! \\
        &&  !let F = zip B E in!\\
        && !foldl (fun (x,y), acc  -> x*acc+y) z1 F)! \\
        && $\nabla_{\Gamma}$!e2! $\nabla_{\Gamma}$!e1,...)!
\end{tabular}
\end{center}

@@

@@

It is slighly better because both derivatives of !e2! and !e3! are put together which might unlock some optimisations, but there is still a non linear usage of the variables.
One solution, if we know we want to compute the whole gradient of the expression, is to define instead

@@

This time, there is linear usage of !Y,z! and the !xi!.

TODO: explain why it's correct?\\
TODO: adding conditionals does not break differentiability as long as no non smooth function like max(0,-):R->B\\
TODO: re explain how it's usually done by partial eval of the conditional first.

@@

We now show how to consider arrays over any ground type $G$.

>>>>

\subsection{Lifting the restriction on reduce}
\label{sub:Lifting the restriction on reduce}

Assume we allow $\Gamma$, !x!: $\reals$, !y!: $\reals$ $\vdash$ !e!: $\reals$ to be the function argument in
!reduce (x,y.e) v A!. We need to add a term depending on $\grad_{\{xi\}}e$ to !yi! in the continuation.
Writing the derivative becomes quite cumbersome, the notation is more compact using arrays of tuples. 
Similarly to he monoid $\widehat{+},0_\Gamma$ defined in Section~\ref{sub:Macro for pure reverse mode transformation},
we use $\widehat{\times},1_\Gamma$ for obvious extension of the monoid $(\reals,\times,1)$.

We can now compute the reverse-mode transformation as follows.
\begin{center}
\begin{tabular}{r c l}
$\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !reduce (x,y.e1) e2 e3!) 
&=& ... as before up until !A3! \\
&& !let B0 = map2 (a,b ->(!$\grad_\Gamma$!e)[a/x,b/y] A0 A) in!\\
&& !let B1 = scanr !$1_\Gamma$ $\widehat{\times}$ !B0 in! \\
&& !let B2 = reduce !$\widehat{+}$ $0_\Gamma$! B1 in! \\
&& !<reduce (x,y.e1) y1 A, fun (x1,...,xn,z) ->! \\
&& !Y((x1,...,xn)!$\widehat{+}$!B2,0,map2 (a,b. a*b*z) A2 A3)!
\end{tabular}
\end{center}

@@

There are two main differences with fold left !foldl! copmpared to !reduce!. 
First, in !foldl (x,y.e) v A! the starting accumulation element !v! is not a unit for !(x,y.e)!,
so it will have non trivial derivatives in general and we need to account for that.
Second, we will need a more general !scanl! which allows !(x,y.e)! as a function argument. 
In other words, we need the general scan left computing the intermediate values of a fold left.
This should be a different primitive but we will still call it !scanl! in this section.

Finally, the former point implies we need a few more array manipulations. 
We write !c::[v1,...,vn]! to be the array ![c,v1,...,vn]!. 
Given an array ![v1,...,vn]! of size $n>0$, we write !let x,A=[v1,...,vn]! to mean 
that !x=v1! and !A=[v2,...,vn]!. The reverse derivative of fold left is then
\begin{center}
\begin{tabular}{r c l}
    $\directD{\rho}{\Gamma}{Y}$($\Gamma\vdash $ !foldl (x,y.e1) e2 e3!) &=&
            !let v,Y1 = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
            && !let A,Y2 = !$\directD{\rho}{\Gamma,v}{Y1}$!(e3) in! \\
            && !let A0 = (scanl (x,y.e1) v A)[1..n] in! \\
            && !let A1 = map2 (a,b.(!$\grad_{\{x\}}$!e1)[a/x,b/y]) A0 A! \\
            && !let A2 = map2 (a,b.(!$\grad_{\{y\}}$!)e1[a/x,b/y]) A0 A! \\
            && !let A3 = scanr * 1 A1! \\
            && !<reduce (x,y.e1) v A, fun (x1,...,xn,z) ->! \\
            && !let y1,B = map2 (x,y. x*y*z) (1::A2) A3 in! \\
            && !Y2(x1,...,xn,y1,B)>! \\
\end{tabular}
\end{center}

TODO: other examples/recipe

@@

It is slightly better because both derivatives of !e2! and !e3! are put together, and this might unlock some optimizations, 
but there is still a non linear usage of the continuation variable !Y!.
One solution, if we know we want to compute the whole gradient of the expression, is to define instead

@@

This time, there is a linear usage of the continuation variable !Y! (and of the variables !x1,...xn! too).

We note that adding conditionals does not break differentiability as long as there are no non-trivial primitives $\reals\to\BB$.
Non smooth functions such as the Rectified Linear Unit (ReLU) in machine learning which can be defined as $reLU(x)\defeq max(0,x)$ 
requires a non smooth primitive such as $>0: \reals\to\BB$. In several AD systems such as TensorFlow, 
the condition is evaluated before differentiation is applied, and so conditionals are never directly differentiated.

@@

We now show how to generalise our reverse-mode transformation to be defined on arrays over any ground type $G$.
That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.
